---
title: "CMNN Burden of Disease Report"
output: html_document
---

# Communicable, Maternal, Neonatal, and Nutritional Disease Burden (CMNN)

```{r}
# Load necessary packages
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(gt)
```


```{r}
# Import the communicable diseases data
data_cmnn <- readr::read_csv("../data/burden-of-disease-cmnn.csv")

# Filter the dataset to include only the three countries chosen by your group.
data_cmnn_mini <- data_cmnn %>%
  filter(Entity %in% c("Botswana", "Brazil", "Brunei"))

# convert column names to lower cases

names(data_cmnn_mini) <- tolower(make.names(names(data_cmnn_mini)))

# rename daly
data_cmnn_mini <- data_cmnn_mini %>%
  rename(
    daly = dalys..disability.adjusted.life.years....communicable..maternal..neonatal..and.nutritional.diseases...sex..both...age..age.standardized..rate.
  )

# rename entity to country
data_cmnn_mini <- data_cmnn_mini %>%
  rename(
    country = entity
  )
```

## Table of Estimates for CMNN Burden Over Time

```{r}
#Widen the table to show trends

wide_data_cmnn_mini <- data_cmnn_mini %>%
  # make sure country and code stay as identifier columns
  pivot_wider(
    id_cols    = c(country, code),
    names_from = year,
    values_from= daly
  ) %>%
  gt() 

wide_data_cmnn_mini
```


## Plot Showing Trends in CMNN Burden Over Time

```{r}
# Here, plot a line graph showing the trend of DALY burden for the chosen countries over time. You can use ggplot2 to create this plot. Each country's line should be a different color.
# NOTE: Use your long 'mini' data frame for plotting, not the wide data used for the table. 

# filter & plot
data_cmnn_mini %>%
  ggplot(aes(x = year, y = daly, color = country)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title    = "DALYs Over Time",
    x        = "Year",
    y        = "Disability-Adjusted Life Years (DALY)",
    color    = "Country"
  ) +
  theme_minimal()

```

## Summary of CMNN Burden Findings

Provide a brief analysis based on the data presented in the table and chart. Highlight any significant findings or patterns. About 3 sentences.

Botswana’s DALYs rose sharply from about 27,900 in 1990 to a peak of roughly 83,200 in 2002 before entering a sustained decline back to about 25,400 by 2019, suggesting a substantial but temporary burden spike in the early 2000s. 

In contrast, Brazil shows a steady and consistent reduction in DALYs from about 13,200 in 1990 down to roughly 4,840 in 2019, indicating long-term improvements in population health. 

Brunei’s DALYs remain comparatively low (around 3,500 in 1990) and exhibit only modest fluctuations over the three decades, reflecting relative stability in its disease burden.

