# Non-Communicable Disease Burden (NCD)

```{r}
# Load necessary packages
pacman::p_load(tidyverse, knitr, here, plotly)

# Import the non-communicable diseases data
data_ncd <- read_csv(here("data/burden-of-disease-ncd.csv"))
```

```{r}
# Filter the dataset to include only the three countries chosen by your group.
data_ncd_mini <- data_ncd %>%
  filter(Entity %in% c("Botswana", "Brazil", "Brunei"))

# convert column names to lower cases

names(data_ncd_mini) <- tolower(make.names(names(data_ncd_mini)))

#rename daly
data_ncd_mini <- data_ncd_mini %>%
  rename(
    daly = dalys..disability.adjusted.life.years....non.communicable.diseases...sex..both...age..age.standardized..rate.
  )

# rename entity to country
data_ncd_mini <- data_ncd_mini %>%
  rename(
    country = entity
  )

```



## Table of Estimates for NCD Burden Over Time

```{r}
# Here render a table for the DALY burden over time for the three countries. 
# You should pivot the data wider to show each country in a separate column.

# Pivot to wide format, keeping country and code as identifiers
data_ncd_wide <- data_ncd_mini %>%
  pivot_wider(
    id_cols    = c(country, code),
    names_from = year,
    values_from= daly
  )

# Render the wide table using kable()
knitr::kable(
  data_ncd_wide,
  caption = "DALYs by Country and Year",
  digits = 2,
  align = c("l", "l", rep("r", ncol(data_ncd_wide) - 2))
)
# Use kable() from the knitr package to render the table.
```

## Plot Showing Trends in NCD Burden Over Time

```{r}
# Here, plot a line graph showing the trend of DALY burden for the chosen countries over time. You can use ggplot2 to create this plot. Each country's line should be a different color.
# NOTE: Use your long 'mini' data frame for plotting, not the wide data used for the table. 
data_ncd_linechart <-data_ncd_mini %>%
  ggplot(aes(x = year, y = daly, color = country)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title    = "DALYs Over Time",
    x        = "Year",
    y        = "Disability-Adjusted Life Years (DALY)",
    color    = "Country"
  ) +
  theme_minimal()

ggplotly(data_ncd_linechart)
```

## Summary of NCD Burden Findings

Provide a brief analysis based on the data presented in the table and chart. Highlight any significant findings or patterns. About 3 sentences.

Between 1990 and 2019, all three countries saw overall declines in non-communicable disease DALYs, with Brazil showing the most consistent year-on-year reduction from about 26 800 to just over 20 000. 

Brunei’s DALYs fell from roughly 29 500 to 22 700, though there was a mid-2000s plateau before renewed improvement. 

In contrast, Botswana experienced a pronounced rise in DALYs—peaking near 32 000 in the mid-2000s—before reversing course to end at approximately 25 600, mirroring regional gains in disease control but highlighting a later start.
